Section_8_Team1_Lab2_Report

### 2021-08-05 notebooks/Draft_Report.Rmd
data description added. pdf formatting.

---
title: "Section_8_Team1_Lab2_Report"
author: "Jonas Degnan, Autumn Rains, Lucy Wu"
date: "8/5/2021"
output:
  html_document:
    df_print: paged
  word_document: default
  pdf_document: default
---

```{r install packages, message=FALSE}
install.packages("table1")
install.packages("GGally")
install.packages("stargazer")
```

``` {r load packages, echo=FALSE}
#we may not need all of these
library(tidyverse)
library(magrittr)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
library(dplyr)
library(table1)
library(GGally)
library(stargazer)
theme_set(theme_minimal())
knitr::opts_chunk$set(echo = TRUE)
#knitr::knit(input = "./../notebooks/Draft_Report.Rmd", output = "./../reports/final_report.pdf")
```

## Introduction

Summer vacations in the United States are synonymous with sunny, leisurely days spent with loved ones. As Ella Fitzgerald so aptly sings, “Summertime, and the living is easy.” However, travel in the recent peak-vacation months has certainly deviated from this idealistic frame of mind with the onset of the COVID-19 pandemic in early 2020. The global travel industry has experienced sharp declines. Specifically in the United States, “travel spending totaled a mere $679 billion in 2020, an unprecedented 42% annual decline (nearly $500 billion) from 2019,” according to analysis from the U.S. Travel Association. 
As infection and vaccination rates oscillate regionally along with government policies for business operation restrictions in the United States, our team of data scientists will look to answer the following research question in more detail using linear regression techniques:

Do COVID-19 infection rates impact vacation travel distance for U.S. residents in the state of Texas? 
Does income, population density, unemployment rates, and age in Texas counties impact traveling?

We will analyze datasets from June, 2021 pertaining to these variables for Texas counties and create three models to understand if causal relationships exist between the number of long distance trips, population income, population density (rural or urban counties). Our hypothesis is that an there is a causal relationship between """

-->Need to explain distance of travel with bs TRANSPORT data
-->Need to add more to causal theory to justify variable selection

## Operationalization

To create our dataframe, we cleaned and filtered data from multiple sources of data and joined based on texas county name or code from June, 2021.

#### Operationalization of focus variables
We gathered covid case and death data from NYT. We gathered trip distance data from the Bureau of Transportation, which they record based on data gathered from personal mobile devices. This data was presented as number of travelers in binned distances. We defined a short trip as less than 100 miles and long trips as greater than 100 miles. This was decided under the assumption that distances traveled of less than 100 miles are for the purpose of commuting, while greater than 100 miles indicated vacation travel. Hand in hand with this variable is our indicator variable for urban vs rural setting *** how did we decide this again, population threshold?**** under the assumption that those who live in rural settings are likely to commute further than those in an urban setting. Finally, we believe that median household income plays a significant role in an individual's ability to take vacations. We pulled this directly from the US Census Bureau.

@@

### Variables

For our analysis, we will use datasets and variables from the four following sources:

[![DataSources](lab2datatable.JPG)](https://ibb.co/zQ3yj2K)

As indicated, each source will provide information on county population demographics such as income, rural vs. urban classification, and status of pandemic regulations. Together, these sources were filtered for the state of Texas and combined to create a dataframe for analysis.

```{r dataframe, echo=FALSE}
df <- read.csv('./../data/processed/covid_trips_processed.csv')
head(df)
```

@@

```{r EDA, echo=FALSE, message=FALSE}
#Example Table
df$rural_cat <- ifelse(df$is_rural == 1, "Rural", "Urban")
table1::label(df$density) <- "County Population Density"
table1::label(df$cases) <- "County Covid-19 Cases"
table1::label(df$unemployed) <- "County Unemployment Rate"
table1::label(df$median_income) <- "County Median Income ($)"
table1::label(df$lt17) <- "County Age < 17 years"
table1::label(df$gt65) <- "County Age > 65 years"
# table1::label(df$is_rural) <- "County Type"
table1::table1(~density + cases + unemployed + median_income + lt17 + gt65 | rural_cat, data = df, caption = "Table 1: Texas County Variable Statistics", topclass="Rtable1-zebra", overall = "Total")
```
Below are histograms of variables that did not follow normal distribution and required transformation. Each of these variables was logarithmically transformed.
```{r scatterplots, message=FALSE}
attach(df)
par(mfrow=c(3,2))
hist(cases, main="Histogram of Covid-19 Cases")
hist(deaths, main="Histogram of Covid-19 Deaths")
hist(long_trip, main="Histogram of Distance")
hist(density, main="Histogram of Population Density")
hist(unemployed, main="Histogram of Unemployment")
hist(median_income, main="Histogram of Median Income")

@@

df %>%
  mutate(
    cases = log(cases+1),
    deaths = log(deaths+1),
    long_trip = log(long_trip),
    density = log(density),
    median_income=log(median_income+1),
    unemployed = log(unemployed+1),
    lt17 = log(lt17),
    gt65 = log(gt65)
  ) %>%
  select(long_trip, cases, deaths, median_income, unemployed, density, lt17, gt65) %>%
  GGally::ggpairs(.,
               legend = 1,
               aes(color = County),
               title = "Figure 1: Variable Characterstics",
               upper = list(continuous = wrap("cor", size = 2.5)),
               lower = list(continuous = wrap("points", alpha = 0.5, size=0.3)),
               diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
               columnLabels=c("Distance", "Cases", "Deaths", "Income", "Unemployment", "Pop Density", "Age < 17","Age > 65"))+
               theme(legend.position = "bottom")

@@

### Accounting Tables

@@

```{r Accounting Table creation & output}
df2 <- df

@@

#### Additional Changes to the Data:

@@

## A Model Building Process

To investigate the causal relationships between distance traveled and variables within the dataset, a linear regression was performed in R Studio for three models. The third model is all-inclusive of the variables available to the team at the time of data gathering. Details of each model analyzed can be found below.

Here is the first model, which is limited to study only key variables of interest for measurement based on our hypotheses. Our key input variable is covid cases and key output is number of trips greater than 100 miles. We also chose to add an indicator for rural counties because it could be directly related to trip distance, as well as median income which plays a role in a person's ability to take vacations in the first place. 
$$
  Model 1 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income})
$$ 
The second model expands upon the first model to also include key explanatory variables and covariates that are hypothesized to improve performance of the previous model. Here, we added in unemployment rate to assess if the impact of covid to that county affected travel.
$$
  Model 2 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income}) + \beta_4 h(\text{Unemployment Rate}) + \beta_5 h(\text{log(Population Density})
$$
Finally in the third model, we added in age demographics as percent under 17/over 65 as another potential source that influences travel midst covid cases due to the nature of how covid affects those at different ages.
$$
  Model 3 = f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income})
  + \beta_4 h(\text{Unemployment Rate}) + \beta_5 h(\text{log(Population Density}) + \beta_6 h(\text{log(% Age < 17}) + \beta_7 h(\text{log(% Age < 17})
$$

## Linear Regression Results

@@

#need to change these to per capita data
model1 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural, data = df) #limited model
model2 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1) + log(density), data = df)
model3 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1) + log(density) + log(lt17) + log(gt65), data = df) #final model
stargazer(model1, model2, model3, type='text', title="Results", align=TRUE)
```
From this output table, we can see most of our hypotheses about the variables we included in our model were correct and do have a statistically significant contribution to our model already.

Additionally, to indicate if any of the models provide a better fit to the dataset than a model that contains no independent variables, an F test was performed. The results for each model are shown below:
```{r F Test}
#Run F test to show statistical significance of additional variables
anova(model1, model2, model3, test = "F")
```
From the F-test,  we can see the the addition of the variables into model 2 and model 3 do contribute significant information to our model based off of the derived p-values.

Last, the significance of individual regression coefficients was tested using robust standard errors in R studio for model 3. The results are shown below: 
```{r coeftest}
#Run coeftest to show statistical significance of all variables in the model with robust standard errors
coeftest(model3, vcov = vcovHC(model3, type="HC1"))
```
From the t-test, we can see that Covid-19 cases and median income are statistically significant. Additionally, the percentage of adults greater than 65, the population density, and the Covid-19 death variables are even more statistically significant.

@@

### Model Limitations

@@

```{r VIF & correlation}

@@

```{r CLM assumptions check}

@@

### Omitted Variable

#### Variable 1

@@

#### Variable 2

@@

## Conclusion & Next Steps

@@

@@

## References

>>>>

---
title: "Section_8_Team1_Lab2_Report"
author: "Jonas Degnan, Autumn Rains, Lucy Wu"
date: "8/5/2021"
output:
  pdf_document: 
    toc: yes
    fig_caption: yes
    fig_crop: no
  word_document: default
  html_document: 
    df_print: kable
    toc: yes
    theme: readable
---
\clearpage
```{r install packages, echo=FALSE, message=FALSE}
# install.packages("table1")
# install.packages("GGally")
# install.packages("stargazer")
# install.packages("tinytex")
```

``` {r load packages, echo=FALSE, message=FALSE}
#we may not need all of these
library(tidyverse)
library(magrittr)
library(ggplot2)
library(patchwork)
library(sandwich)
library(lmtest)
library(dplyr)
library(table1)
library(GGally)
library(stargazer)
theme_set(theme_minimal())
options(tinytex.verbose = TRUE)
knitr::opts_chunk$set(echo = TRUE)
# tinytex::install_tinytex()
#knitr::knit(input = "./../notebooks/Draft_Report.Rmd", output = "./../reports/final_report.pdf")
```

# Introduction

Summer vacations in the United States are synonymous with sunny, leisurely days spent with loved ones. As Ella Fitzgerald so aptly sings, “Summertime, and the living is easy.” However, travel in the recent peak-vacation months has certainly deviated from this idealistic frame of mind with the onset of the SARS-CoV-2 outbreak (COVID-19) in early 2020. The global travel industry has experienced sharp declines. Specifically in the United States (US), “travel spending totaled a mere $679 billion in 2020, an unprecedented 42% annual decline (nearly $500 billion) from 2019,” according to analysis from the U.S. Travel Association. 
As infection and vaccination rates oscillate regionally along with government policies for business operation restrictions in the United States, our team of data scientists will look to answer the following research question in more detail using linear regression techniques:

Do COVID-19 infection rates impact vacation travel distance for U.S. residents in the state of Texas? 
How does income, population density, and covid regulations in each county impact traveling?

We will analyze data pertaining to these variables for Texas counties and create three models to understand if causal relationships exist between the number of long distance trips, population income, population density (rural or urban counties). Our hypothesis is that an there is a causal relationship between """ [summer travel]

-->Need to explain distance of travel with bs TRANSPORT data[^bts01] [survey](https://www.bts.gov/statistical-products/surveys/national-household-travel-survey-long-distance-travel-quick-facts)
-->Need to add more to causal theory to justify variable selection

**Causal theory needed to identify why these particular variables below were chosen**

# Data Description

Operationalization of our research question required data from four sources: the New York Times (NYT)' _Coronavirus (Covid-19) Data in the United States_ dataset[^nyt], the United States Census Bureau's 2010 decennial survey[^uscb], the Texas Association of Counties' _County Information Program_[^tac] database, and the United States Bureau of Transportation Statistics' _Trips by Distance_[^bts] dataset.

The NYT has maintained a robust public dataset of county-level, time-series COVID data on GitHub and holds records dating back to the beginning of COVID-19. The dataset aggregates state and municipal government health department reports of cumulative cases and deaths for each US county daily from COVID-19. NYT data does not match deaths or cases for patients to their home state (e.g., vacationers), nor does it identify if a case occurred in one county, but a subsequent death occurred in another. Therefore, there is a likely mismatch between this data and those reported by individual states and counties.

For our investigation, we selected on cases and deaths in the month of June 2021, to best operationalize summer. The case and death counts were aggregated by county to mitigate time-series affects on modeling. The death counts were identified as being directly related to the COVID-19 infection and not from other instances (e.g., homicide), or unrelated comorbidities (e.g., cancer).

The Texas Association of Counties proved to be a rich source of demographic and economic data for our target population. This database aggregates national and state data sources (e.g., Census Bureau, Texas Workforce Commission, United States Bureau of Labor Statistics, United States Bureau of Economic Analysis, and the Texas Demographic Center). The lineage of the TAC data was challenging to individually trace. All of the data originated from primary sources at the Federal or State level, but there was no readily available documentation to map columnar data back to its primary source. We selected five measurements to support our causal theory: county population and density, median household income, number of individuals in poverty, and the unemployment rate. The economic measurements of poverty and employment were hypothesized to have an inverse relationship with enabling people to take long vacation trips, where median income should have a more direct relationship. The county population served as a control for the economic variables and density was hypothesized to encourage longer trips to more open spaces.

The decennial survey conducted by the Census results in an incredibly rich national database of economic, geographic, demographic, and social measures. While the vast database has many of the measurements useful to support our causal theory, as the TAC database queries held a majority of these, the decennial survey did proffer a means to generate an indicator variable identifying rural or urban counties. This variable was important since we hypothesized that rural counties may experience higher long trip counts due to their geographies.

Our final course were data produced by the US Department of Transportation's Bureau of Transportation Statistics (USBTS). These data are experimental *estimates* of mobility gathrered from aggregated mobile device data. USBTS states that these data are experimental, and but significant effort was made to ensure proper population weighting, and data multi-sourcing (to reduce variance in spacial and temporal measurements common from a single source) were implemented before any of the statistics were calculated. Trips are defined as a longer than 10 minute stop at a location that is not home. Trips are not distiguish by mode of transporation (i.e., air, train, bus, car, or bike). For our analyis, we were interested in only trips longer than 100 miles in Texas in June of 2021 as this best operationalized the concept of a vacation[^bts01].

[^bts01]: https://www.bts.gov/statistical-products/surveys/national-household-travel-survey-long-distance-travel-quick-facts
[^nyt]: The New York Times. (2021, 08 04). Coronavirus (Covid-19) Data in the United States. https://github.com/nytimes/covid-19-data. Retrieved 07 21, 2021, from https://raw.githubusercontent.com/nytimes/covid-19-data/master/us-counties.csv
[^uscb]: United States Census Bureau. (2011, 11 22). Decennial Census, 2010. Explore Census Data. Retrieved 07 21, 2021, from https://data.census.gov/
[^tac]: Texas Association of Counties. (2021, 08 04). QueriesCIP. County Information Program. Retrieved 08 02, 2021, from https://imis.county.org/iMIS/CountyInformationProgram/
[^bts]: Department of Transportation. (2021, 08 02). Trips by Distance. Bureau of Transportation Statistics. Retrieved 07 19, 2021, from https://data.bts.gov/Research-and-Statistics/Trips-by-Distance/w96p-f2qv

---
#### Operationalization of focus variables

__integrating below still__
We gathered covid case and death data from NYT. We gathered trip distance data from the Bureau of Transportation, which they record based on data gathered from personal mobile devices. This data was presented as number of travelers in binned distances. We defined a short trip as less than 100 miles and long trips as greater than 100 miles. This was decided under the assumption that distances traveled of less than 100 miles are for the purpose of commuting, while greater than 100 miles indicated vacation travel. Hand in hand with this variable is our indicator variable for urban vs rural setting *** how did we decide this again, population threshold?**** under the assumption that those who live in rural settings are likely to commute further than those in an urban setting. Finally, we believe that median household income plays a significant role in an individual's ability to take vacations. We pulled this directly from the US Census Bureau.

@@

As indicated, each source will provide information on county population demographics such as income, rural vs. urban classification, and status of pandemic regulations. Together, these sources were filtered for the state of Texas and combined to create a dataframe for analysis.
---

```{r dataframe, echo=FALSE}
df <- read.csv('./../data/processed/covid_trips_processed.csv')
# head(df)
```

@@

```{r EDA, echo=FALSE, message=FALSE}
#Example Table
df$rural_cat <- ifelse(df$is_rural == 1, "Rural", "Urban")
table1::label(df$density) <- "County Population Density"
table1::label(df$cases) <- "County Covid-19 Cases"
table1::label(df$unemployed) <- "County Unemployment Rate"
table1::label(df$median_income) <- "County Median Income ($)"
table1::label(df$lt17) <- "County Age < 17 years (%)"
table1::label(df$gt65) <- "County Age > 65 years (%)"
# table1::label(df$is_rural) <- "County Type"
table1::table1(~density + cases + unemployed + median_income + lt17 + gt65 | rural_cat, data = df, caption = "Table 1: Texas County Variable Statistics", topclass="Rtable1-zebra", overall = "Total")
```
Below are histograms of variables that did not follow normal distribution and required transformation. Each of these variables was logarithmically transformed.
```{r scatterplots, message=FALSE, echo=FALSE}
attach(df)
par(mfrow=c(3,2))
hist(cases, main="Histogram of Covid-19 Cases")
hist(deaths, main="Histogram of Covid-19 Deaths")
hist(long_trip, main="Histogram of Distance")
hist(density, main="Histogram of Population Density")
hist(unemployed, main="Histogram of Unemployment Rate")
hist(median_income, main="Histogram of Median Income")

@@

df %>%
  mutate(
    cases = log(cases+1),
    deaths = log(deaths+1),
    long_trip = log(long_trip),
    density = log(density),
    median_income=log(median_income+1),
    unemployed = log(unemployed+1)
  ) %>%
  select(long_trip, cases, deaths, median_income, unemployed, density, lt17, gt65) %>%
  GGally::ggpairs(.,
               legend = 1,
               aes(color = County),
               title = "Figure 1: Variable Characterstics",
               upper = list(continuous = wrap("cor", size = 2.5)),
               lower = list(continuous = wrap("points", alpha = 0.5, size=0.3)),
               diag = list(continuous = wrap("densityDiag", alpha = 0.5)),
               columnLabels=c("Distance", "Cases", "Deaths", "Income", "Unemployment", "Pop Density", "Age < 17","Age > 65"))+
               theme(legend.position = "bottom")

@@

## Accounting Tables

@@

```{r Accounting Table creation & output, echo=FALSE}
df2 <- df

@@

## Additional Changes to the Data:

@@

# A Model Building Process

To investigate the causal relationships between distance traveled and variables within the dataset, a linear regression was performed in R Studio for three models. The first model is limited to study only key variables of interest for measurement. The second model expands upon the first model to also include key explanatory variables and covariates that are hypothesized to improve performance of the previous model. The third model is all-inclusive of the variables available to the team at the time of data gathering. Details of each model analyzed can be found below.

$Model_1: f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income})$

$Model_2: f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income}) + \beta_4 h(\text{Unemployment Rate}) + \beta_5 h(\text{log(Population Density})$

$Model_3: f(\text{log(Number of Long Trips)}) = \beta_0 + \beta_1 g(\text{log(Covid-19 Cases)})  + \beta_2 h(\text{Rural County}) + \beta_3 h(\text{Median Income}) + \beta_4 h(\text{Unemploymed}) + \beta_5 h(\text{log(Population Density}) + \beta_6 h(\text{log(Age < 17}) + \beta_7 h(\text{log(Age > 65}) + \beta_7 h(\text{log(In Poverty})$

# Linear Regression Results

@@

#need to change these to per capita data
model1 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + log(median_income+1), data = df) #limited model
model2 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + median_income + unemployed + log(density), data = df)
model3 <- lm(log(long_trip) ~ log(cases+1) + log(deaths+1) + is_rural + median_income + unemployed + log(density) + lt17 + gt65, data = df) #final model
stargazer(model1, model2, model3, type='text', title="Results", align=TRUE)
```
Additionally, to indicate if any of the models provide a better fit to the dataset than a model that contains no independent variables, an F test was performed. The results for each model are shown below:
```{r F Test, echo=FALSE}
#Run F test to show statistical significance of additional variables
anova(model1, model2, model3, test = "F")
```
From the F-test,  we can see the the addition of the variables into model 2 and model 3 do contribute signficant information to our model based off of the derived p-values.

Last, the significance of individual regression coefficients was tested using robust standard errors in R studio for model 3. The results are shown below:
```{r coeftest, echo=FALSE}
#Run coeftest to show statistical significance of all variables in the model with robust standard errors
coeftest(model2, vcov = vcovHC(model3, type="HC1"))
```
From the t-test, we can see that Covid-19 cases and median income are statistically significant. Additionally, the percentage of adults greather than 65, the population density, and the Covid-19 death variables are even more statistically significant.

@@

# Model Limitations

@@

```{r VIF & correlation, echo=FALSE}

@@

```{r CLM assumptions check, echo=FALSE}

@@

# Omitted Variable

### Vehicle Ownership

@@

### Fuel Prices

@@

# Conclusion

@@

### Next Steps

@@
